---
editor_options:
  markdown:
    wrap: 72
---

# An Introduction to `lavaan`: How Do Wet Fingers Wrinkle? {.unnumbered}

:::file
For this demonstration, we will make use of data from the following study:

Kareklas, et al. (2013) "Water-induced finger wrinkles improve handling of wet objects", Biology Letters,
<http://dx.doi.org/10.1098/rsbl.2012.0999>
:::

## Workspace {.unnumbered}

First, we're going to clear the workspace (beware! This will remove
everything from the "environment").

```{r}
rm(list = ls())
```

Now, let's set the working directory. You will remember that the working
directory is the folder where R will look for and save files to. If you
want to know what your current working directory is, run the command
below:

```{r eval = FALSE}
getwd() 
```

To change it, modify the path in the quotation marks `" "` below to your
directory.

```{r eval = FALSE}
setwd("C:/Users/...")
```

Now let's load the relevant packages. In this demonstration, we'll be
using the `lavaan` library, but also a few others: `fastDummies` which
will facilitate the generation of dummy variables, and `emeans` for
computing and plotting estimated marginal means. You must install these
prior to loading them.

```{r eval = FALSE}
install.packages("lavaan")
install.packages("fastDummies")
install.packages("emmeans")
```

In addition to these, we will also need `tidyverse` to allow us to make
use of the pipe operator, `ggplot2` for plotting results and `janitor`
for generating frequency tables.

```{r warning = FALSE, message = FALSE}
library("lavaan")
library("tidyverse")           
library("janitor")             
library("fastDummies")        
library("ggplot2")             
library("emmeans")             
```

## Data {.unnumbered}

We'll download the data from the internet, and store it in a dataframe
called **wrinkle**.

```{r}
wrinkle <- read.csv("https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Wrinkle.csv")
```

## Experimental Design {.unnumbered}

The skin on our fingers and toes wrinkles when it has been in water for
a few minutes. Why on earth does it do this? The skin on almost all
other body parts does not wrinkle when wet.

Twenty participants were asked to pick up small objects with the right
hand thumb and index finger, pass them through a small hole, grab them
with the left hand, and put them into a box that had a hole in the lid.
For half the trials, participants had their fingers soaked in water,
making the skin wrinkle. On the other half of the trials, the fingers
were dry. For half the trials the objects were wet. For the other half,
the objects were dry.

The time it took the participants to finish the task of passing all of
the objects into the box was recorded for each condition:

-   Condition 1: Wet/wrinkled fingers, wet objects\
-   Condition 2: Wet/wrinkled fingers, dry objects\
-   Condition 3: Dry fingers, wet objects\
-   Condition 4: Dry fingers, dry objects

*Hypothesis*: Skin wrinkling is an adaptation to improve grip when in
wet environments. That's why we only need it on our fingers (for
handling wet objects) and toes (for walking on wet surfaces).

*Research Question*: Will objects be easier/quicker to handle with
wrinkled/wet fingers compared to with dry fingers?

## Frequency for Variables and Levels {.unnumbered}

Fingers were wet/wrinkled half the time

```{r}
wrinkle %>% tabyl(Fingers) 
```

Objects were wet half the time

```{r}
wrinkle %>% tabyl(Objects)
```

Participants 1 to 20 (all four conditions)

```{r}
wrinkle %>% tabyl(Participant) 
```

## The Causal Model {.unnumbered}  

This was an experiment. The values of the predictors (**Fingers** and
**Objects**) were decided by the experimenters, so they are uncorrelated
causes of the outcome (**Time**).

**Specific Hypotheses:**

1.  Handling wet objects will take longer than handling dry objects  

2.  Handling objects with wet fingers will take longer than with dry fingers  

3.  Handling wet objects will be faster if done with wet/wrinkled fingers than with dry fingers   

4.  Handling time for dry objects will NOT be affected by whether the fingers are wet or dry  


## Statistical Model {-}

The hypothesis is not simply that wet fingers and wet objects will
result in slower handling time, it is that wet fingers will reduce the
slowing effect of handling wet objects specifically. This implies an
interaction between **Fingers** and **Objects**.

## Preparing the Data {.unnumbered}

First we'll create numerical dummy variables for having wet (wrinkled)
fingers and wet objects.

```{r}
wrinkle <- dummy_cols(wrinkle, select_columns = "Fingers")
wrinkle <- dummy_cols(wrinkle, select_columns = "Objects")
```

We now have two new dummy variables: **Fingers_wrinkled** and
**Objects_wet**

```{r}
head(wrinkle)
```

And then we create an interaction variable for **Fingers** and
**Objects**.

```{r}
wrinkle$Interaction <- wrinkle$Fingers_wrinkled*wrinkle$Objects_wet
```

Finally, lets's re-organise the data so that it can be used with
`ggplot2`, so we can visualise the data. We need to convert the
**Participant** variable to a numeric variable so that we can sort it
correctly. We'll use the `str_replace()` function from the `stringr`
library to remove the "p" from the **Participant** variable, and then
convert it to a numeric variable using `as.numeric()`. We'll then use
`relocate()` from `dplyr` to move the **id** variable to the second
column. Finally, we'll use `arrange()` also from `dplyr` to sort the
dataframe by the **id** variable. We'll save the sorted dataframe as
**wrinkle2**.

```{r}
wrinkle2  <- wrinkle %>% 
  mutate(id = as.numeric(str_replace(Participant, "p", ""))) %>%
  relocate(id, .after = Participant) %>%
  arrange(id)
```

## Visualing the Data {.unnumbered}

Great, now we have a dataset that we can use with `ggplot2`. We can use
either **wrinkle** or **wrinkle2** to analyse the data with `lm()` and
`lavaan`.

Let's see the completion time for each condition as a boxplot.

```{r}
wrinkle2 %>%
  ggplot(aes(x = Fingers, y = Time)) +
  geom_boxplot() +
  facet_wrap( ~ Objects)
```

We have four boxplots, one for each condition. The thick black
horizontal line in the middle of each box is the median time to complete
the task. The upper and lower edges of the box are the 75th and 25th
percentiles of completion times for each condition. The "whiskers"
extend to the most extreme completion times, that represent the 2.5% and
97.5% percentiles of the data. The dots are "outliers" - completion
times that are so extreme that they are outside of the whiskers, which
represent 95% of the data. It's clear that handling wet objects with dry
("non" wrinkled) fingers has the highest median handling time.

Let's now take a look at a line graph of the interaction. First, we
compute a summary statistic (the mean) for each condition.

```{r}
wrinkle2_summary  <- wrinkle2 %>%
  group_by(Fingers, Objects) %>%
  summarise(Mean_time = mean(Time))
```

Then, we plot the above statistics.

```{r}
wrinkle2_summary %>%
  ggplot(aes(x = Fingers, y = Mean_time, color = Objects)) +
  geom_line(aes(group = Objects)) +
  geom_point()
```

The dots show the mean completion time for each condition. The lines
connect the dots for the same object type. The lines are not parallel,
which suggests that we have an interaction between Fingers and Objects.
However, this graph does not show us the uncertainty in the mean
completion times. (The graph is also misleading because the y-axis does
not begin at zero, which makes the differences across contitions look
larger than they are.)

To fully assess our hypotheses, we need to fit statistical models of the
mean differences, which take into account the uncertainty in the mean
completion times, by computing standard errors (SEs) for the
differences.   


## The `lavaan` Package {-}  

With `lavaan`, we define the statistical model as an object and we name it **model**. The object specification is enclosed in single quotes. As you can see below, the model structure is similar to that of `lm()`. 

```{r}
model <- 'Time ~ Fingers_wrinkled + Objects_wet + Interaction'
```

To then fit the model to the data, we use the `sem()` function. 

```{r warning = FALSE, message = FALSE}
fit_lav <- sem(model, data=wrinkle2, meanstructure = TRUE)
```

One difference is that `lavaan` does not include intercepts in the model by default, so we have to ask for them by including the option `meanstructure = TRUE`. 

Let's take a look at the output.

```{r}
summary(fit_lav)
```

As you can see, there is a lot more output than from `lm()`! For now, we will focus on the part starting at `regressions`.

## Tasks and Questions {-}

### Task 1 {-}

Fit a linear model using `lm()` to evaluate the effect of the predictors **Fingers** and **Objects**, and their interaction, on outcome **Time**.

Are the specific hypotheses supported by the data?   


### Task 2 {-}

What are the differences between the `lm()` and `lavaan` results? Why might these differences exist?  


### Task 3 {-}

How have the results changed after relaxing the independence assumption?
Would your answer to the research question change?  

